---
title: "PS270 Final Project"
author: "Decker Storch"
description: "Congress Demographics Over the 20th and 21st Centuries"
output:
  distill::distill_article:
    self_contained: false
---

## Introduction

What is the relationship between the general composition of Congress members and their party affiliation over time? This project looks at the dynamic reflection of generational forces within the composition of Congress to investigate the relationship between the generational composition of Congress and member's party affiliation over time. I hypothesize that members of younger generations would make up a more significant portion of the Democratic portion of Congress, while the older members would make up more of the Republican portion. This is due to the generational theory that younger people are often associated with more progressive ideals due to differences in social and political values. Therefore, during Congresses with lower average ages among members, I expect a strong democratic percentage and in Congresses with older average ages, I expect a more conservative dominant Congress.

Understanding these trends are important to explore as we look to understand generational shifts and how they can influence political participation and party affiliation in Congress. It can have practical implications on highlighting disparities in representation and predicting legislative priorities by looking at historical and current patterns that can contribute to ideas surrounding partisan realignment and generational behavior. For my hypothesis to be supported, I would have to see a higher proportion of the youngest generations in the data, specifically Millennials and Gen Z, to be associated with the democratic party compared to the older generations in the dataset. Similarly, I would like to see an observable difference in party affiliation between younger and older generations in Congress. My hypothesis would be disproved if party affiliation is relatively evenly distributed across generations within Congress or if the younger generations turn out to have a stronger affiliation with the Republican party compared to the older generations, or vice versa. Disproving the hypothesis would challenge my presumptions about generation alignment with political ideology, which would offer new insight into complexities of political identity.

## Data Visualization and Analysis

This project uses a Congress Demographics dataset compiled and organized by FiveThirtyEight. The dataset includes various variables such as generations, birth years, and party affiliation of Congressional members from the 66th to the 118th Congress, which pertains to the years 1919-2023. My explanatory variable of interest is the generation to which each congress member belongs based on their year of birth. This is measured by assigned names to generations defined by year ranges, such as Boomers, the Progressive Generation, Millennials, the Lost Generation, etc. My outcome variable of interest is the party affiliation of any given senator, which is measured by a code that assigns a value to party affiliation. For example, in the dataset democrats are assigned the party_code = 100, republicans = 200, and independents = 328. Additional codes capture less common affiliations that are reflected in my charts.

The research design for my project is longitudinal cross-sectional as I analyze data from many Congresses, treating each one individually to analyze longitudinal trends over the whole data. Through this approach, like I mentioned in the introduction, I can analyze both individual Congress' generation and party affiliation percentages while also analyzing historical trends. To visualize the base distribution of party affiliations across generations, I created a bar chart showing the proportion of Congressional members of each generation grouped by their individual party affiliation. Through analyzing the chart we can see basic trends that can help us later on such as the significant Democratic presence in older generations like the Lost and Silent Generations, while we can see a much more balanced representation across party lines, but also a significantly smaller sample size of younger people.

```{r, include = FALSE}
library(tidyverse)
library(readr)
library(ggplot2)
library(infer)
congress_demographics <- read_csv("congress-demographics/data_aging_congress.csv")
head(congress_demographics)
```

```{r, fig.width = 12, fig.height = 7}
congress_counts <- congress_demographics |>
  mutate(party = case_when(party_code == 100 ~ "Democrat",
                           party_code == 200 ~ "Republican",
                           party_code == 328 ~ "Independent",
                           party_code %in% c(537, 523, 522, 402, 380, 370, 356, 347, 331, 329, 112) ~ "Other"))

congress_counts <- congress_counts |>
  mutate(party = factor(party, levels= c("Democrat", "Republican", "Independent", "Other")))

congress_counts <- congress_counts |>
  count(generation, party) |>
  rename(count = n)

congress_counts <- congress_counts |>
  mutate(generation = factor(generation, levels = c(
    "Gilded", "Progressive", "Missionary", "Lost", "Greatest", "Silent", "Boomers", "Gen X", "Millennial", "Gen Z")))

congress_counts |>
  ggplot(mapping = aes(x = generation, y = count, fill = party)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Democrat" = "blue", "Republican" = "red", "Independent" = "purple1", "Other" = "grey")) +
  labs(x = "Generation of Congress Member",
       y = "Proportion of Members by Party",
       fill = "Party Affiliation",
       title = "Party Affiliation Across Generations in Congress",
       caption = "Source: Congress Demographics Dataset for FiveThirtyEight") +
  theme_minimal()
```

## Results

```{r, fig.width = 12, fig.height = 7}
generation_percentages <- congress_demographics |>
  group_by(congress, generation) |>
  summarise(count = n(), .groups = "drop") |>
  group_by(congress) |>
  mutate(percentage = (count / sum(count)) * 100)


generation_percentages |>
  ggplot(mapping = aes(x = congress, y = percentage, color = generation)) +
  geom_line(size = 1) +
  geom_point() +
  scale_color_manual(values = c("Gilded" = "orange", "Progressive" = "cyan",
                                "Missionary" = "green", "Lost" = "blue",
                                "Greatest" = "navy", "Silent" = "purple",
                                "Boomers" = "pink", "Gen X" = "forestgreen",
                                "Millennial" = "red", "Gen Z" = "yellow3")) +
  labs(x = "Congress Number",
       y = "Percentage of Members",
       color = "Generation",
       title = "Generational Composition of Congress Over Time: 66th to 118th",
       caption = "Source: Congress Demographics Dataset for FiveThirtyEight") +
  theme_minimal()
```

```{r}
congress_reg_demographics <- congress_demographics |>
  mutate(party_code_binary = case_when(party_code == 100 ~ 1, party_code == 200 ~ 0, TRUE ~ NA_real_))

congress_reg_demographics <- congress_reg_demographics |>
  filter(!is.na(party_code_binary))

model <- glm(party_code_binary ~ age_years, data = congress_reg_demographics, family = binomial)

summary(model)$coefficients
```

Upon observation of both graphs, an obvious issue with my hypothesis is that due to general trend of higher average ages of Congress members, there are not many younger representatives in the data for us to analyze trends and potentially understand the patterns of liberal or conservative ideology among younger representatives. Through analysis of the first barchart we created we can develop an understanding about the percentages of Democrats and Republicans by generation (age). This understanding allows us to then make educated inferences about Congresses age demographics that are displayed in the second graph, which shows us the various percentages of certain generations within each individual Congress. However, building a regression model for this was significantly tougher than I originally thought, as there are so many pieces affecting the coefficients. So I narrowed my regression by removing Independent and Other party affiliations, as they are a significant minority in all Congresses.

Therefore, after creating a general linear regression model of the large quantity of data over time, the data returned a coefficient of -0.0008087, meaning that for each additional year of age of a Congress member, Democratic affiliation is decreasing by less than 0.1 percentage points. The summary also returns a p-value, shown by the Pr(>|z|), of 0.46, which is well above the standard 0.05 we look for to reject the null hypothesis and determine that the relationship is not statistically significant. I was expecting the coefficient to be small due to the large amount of data points and, p-value aside, the observed negative value of the coefficient supports my stated hypothesis that older generations trend towards the Republican party, even if it is by such a small amount. However, due to the p-value being 0.46, we cannot interpret the relationship between age and Republican identification causally as there is too big a risk for other confounding variables that could be affecting the data, suggesting that the observed trend is likely due to chance rather than a true relationship between age and party affiliation.

## Conclusion

In conclusion, upon looking at the proportion of party members and then comparing that to the percentage of by members by generation within each recorded congress, we can see that the results of this analysis provide very limited support for my hypothesis. While the regression showed a small negative coefficient, indicating a very slight association between older age and democratic affiliation, the p-value of 0.46 suggests that the relationship is far from statistically significant. I believe this lack of significance stems from certain limitations of the dataset stemming from small sample sizes of the youngest eligible generations of Millenials and Gen X along with other confounding variables such as historical events, generational shifts, and other aspects that would affect party affiliation beyond simply the age of the representative. So, if I had more time and money, I would focus my regressions to control and focus confounding variables, such as running multiple regressions for certain time periods or regions of the United States.

